---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Angel Rutherford"
date: "2025-09-28"
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Pennsylvania Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}


# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("595f8315942e8965e5a91f7cfa62b7210fe62245")
# Choose your state for analysis - assign it to a variable called my_state
my_state <- "PA"
```

**State Selection:** I have chosen **Pennsylvania** for this analysis for its personal relevance since I currently live here.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
county_data <- get_acs(
  geography ="county",
  variables= c(
     median_income ="B19013_001",
     total_pop="B01003_001"
  ),
  state = my_state,
  year = 2022,
  output = "wide"
)
# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
county_data <- county_data%>%
  mutate(county_name = str_remove(NAME, " County, Pennsylvania"))
         
# Display the first few rows
glimpse(county_data)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
county_data <-county_data%>%
  mutate(MOE_PCT=(median_incomeM/median_incomeE)*100,
  reliability = case_when(
    MOE_PCT < 5 ~ "High Confidence (<5%)",
    MOE_PCT <= 10 ~ "Moderate Confidence (5%-10%)",
    TRUE ~ "Low Confidence (>10%)"
    )
  )
county_data
# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
reliability_summary <- county_data%>%
  count(reliability) %>%
  mutate(
    percentage = round(n/sum(n) * 100, 1),
    total = sum(n)
  )
reliability_summary
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.
```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
top_5_counties <- county_data %>%
  arrange(desc(MOE_PCT))%>%
  slice_head(n=5)

# Format as table with kable() - include appropriate column names and caption 
top_5_counties %>%
  select(county_name, median_incomeE, median_incomeM, MOE_PCT, reliability)%>%
    kable(
      col.names = c("County Name", "Income ($)", "MOE", "MOE Percentage (%)", "MOE Reliability"),
      caption = "Top 5 Counties Exhibiting High Uncertainty",
     digits=c(0,0,1,1,0)
)
  
```

**Data Quality Commentary:**

These results highlight that although all five counties are all considered to indicate moderate confidence in their income estimations, their margins of error vary significantly. For instance, the county with the largest margin of error, Forest County, lies on the cusp of exhibiting low confidence while the smallest margin of error, Elk County, is much closer to being high confidence than low. These discrepancies, in addition to general uncertainty, indicate a need for additional oversight when using income data from these counties for decision-making.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties <- county_data %>% 
  filter(county_name %in% c("Allegheny", "Carbon", "Forest"))   %>%
  select(county_name, median_incomeE, MOE_PCT, reliability)

selected_counties
```

**Comment on the output:** Because my data possessed no findings with low confidence, I chose the top county from my "Top 5 Counties Exhibiting High Uncertainty" as my low confidence observation. A gradient in reliability, however, is still achieved as, as previously observed, the margins of error between both counties with estimations of moderate confidence range drastically.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
find_codes <- county_data %>% 
  select(county_name, GEOID)%>%
  filter(county_name %in% c("Allegheny", "Carbon", "Forest"))   %>%
  mutate(county_code = str_remove(GEOID, "42"))
find_codes

tract_data <- get_acs(
  geography ="tract",
  variables= c(
     white ="B03002_003",
     black="B03002_004",
     hispanic="B03002_012",
     total_pop="B03002_001"
  ),
  state = my_state,
  county =c(003, 025, 053),
  year = 2022,
  output = "wide"
)

tract_data
  
# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
# Add readable tract and county name columns using str_extract() or similar

unreadable= str_split(tract_data$NAME, ";", simplify =  TRUE)

tract_data <-tract_data%>% 
  mutate(white_pct=(whiteE/total_popE)*100,
         black_pct=(blackE/total_popE)*100,
         hispanic_pct=(hispanicE/total_popE)*100,
         tract_name=unreadable[,1],
         county_name=str_remove(unreadable[,2], "County")
        
)

glimpse(tract_data)
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
highest_hispanic_pct <- tract_data %>%
  arrange(desc(hispanic_pct))%>%
  slice(1)
glimpse(highest_hispanic_pct)
# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
summary_by_county <- tract_data %>%
  group_by(county_name) %>%
  summarize(
    number_of_tracts=n(),
    mean_white_pct=round(mean(white_pct, na.rm = TRUE),1),
    mean_black_pct=round(mean(black_pct, na.rm = TRUE),1),
    mean_hispanic_pct=round(mean(hispanic_pct, na.rm = TRUE),1)
  )
summary_by_county
# Create a nicely formatted table of your results using kable()
summary_by_county%>%
 kable(
      col.names = c("County Name", "Total Tracts Per County", "Average of White Residents(%)", "Average of Black Residents(%)", "Average of Hispanic Residents (%)"),
      caption = "Average Demographics Per County",
     digits=c(0,0,1,1,1)
)
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
tract_data <-tract_data%>%
  mutate(
    white_moe_pct=(whiteM/whiteE)*100,
    black_moe_pct=(blackM/blackE)*100,
    hispanic_moe_pct=(hispanicM/hispanicE)*100,
    moe_flag=ifelse(
    white_moe_pct > 15 |
    black_moe_pct > 15 |
    hispanic_moe_pct > 15,
    TRUE, FALSE)
)

# Create summary statistics showing how many tracts have data quality issues
 quality_summary <- tract_data%>%
  count(moe_flag) %>%
  mutate(
    percentage = round(n/sum(n) * 100, 1),
    total = sum(n)
  )
quality_summary
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages

# Use group_by() and summarize() to create this comparison
summary_by_moe <- tract_data %>%
  group_by(moe_flag) %>%
  summarize(
    number_of_tracts= n(),
    mean_population=round(mean(total_popE, na.rm = TRUE),0),
    mean_white_pct=round(mean(white_pct, na.rm = TRUE),1),
    mean_black_pct=round(mean(black_pct, na.rm = TRUE),1),
    mean_hispanic_pct=round(mean(hispanic_pct, na.rm = TRUE),1)
  )
summary_by_moe
# Create a professional table showing the patterns
summary_by_moe%>%
 kable(
      col.names = c("Quality Issues (MOE > %15)", "Total Tracts Detected", "Average Total Population", "Average of White Residents(%)", "Average of Black Residents(%)", "Average of Hispanic Residents (%)"),
      caption = "Average Demographics Per County",
     digits=c(0,0,0,1,1,1)
)
```

**Pattern Analysis:** All of the tracts investigated possessed high margins of error and predominantly white communities. This may suggest that demographic data from less diverse communities is more likely to be unreliable.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

**Overall Pattern Identification**
Across all analyses, the smaller the community being surveyed, the less accurate the data would become. Although the counties that were chosen to be further analyzed at the tract level represented a gradient in income estimation reliability, they all shared similar unreliability in demographic data. Allegheny, the county with the smallest margin of error in income estimation, possessed the largest population of the chosen counties while Forest county, the county with the largest margin of error, possessed the smallest population. These counties similarly, however, possessed a disproportionately large number of white residents compare to other racial and ethnic groups and returned high margins of error in their demographic data.

**Equity Assessment**
The county data here emphasizes that margins of error vary depending on the type of data collected and are influenced by both the absolute size of a population and its proportion within the broader population. More specifically, smaller groups such as counties with small populations and racial or ethnic minorities are more prone to data inaccuracies and thus more vulnerable to bias in algorithmic decision-making. For instance, the tract with the largest hispanic population, the smallest average population across all tracts, had an even larger margin of error. Decisions made based on the this demographic data may vastly misrepresent the true needs of the hispanic population.

**Root Cause Analysis**
The misrepresentation of vulnerable populations such as small or spatially uneven populations in data can result in algorithms using inaccurate inputs and thus, inherently reproducing non-representational outputs. Furthermore, the subjective choices made when analyzing data can also skew outputs.

**Strategic Recommendations**
To mitigate these risks, the Department should identify margins of error and create thorough thresholds for interpretations of margins of errors. Certain thresholds should call for human oversight and require additional social, historical, and spatial explanations. 

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()
county_summary <-county_data%>%
  mutate(
    rec_for_algorithm= case_when(
    reliability == "High Confidence (<5%)" ~ "Safe for algorithmic decisions",
   reliability == "Moderate Confidence (5%-10%)" ~ "Use with caution - monitor outcomes",
    reliability == "Low Confidence (>10%)" ~ "Requires manual review or additional data"
    )
  )%>%
  select(county_name, median_incomeE, MOE_PCT, reliability, rec_for_algorithm)%>%
  arrange(desc(MOE_PCT))%>%
  kable(
    col.names = c("County Name", "Median Income ($)", "MOE Percentage(%)", "Reliability", "Algorithm Recommendation)"),
      caption = "County-level Algorithmic Recommendations",
     digits=c(0,0,1,0,0)
  )
county_summary
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** 
Allegheny, Montgomery, Bucks, Chester, Delaware, Lancaster, York, Northampton, Lehigh, Cumberland, Dauphin, Beaver, Luzerne, Washington, Schuylkill, Erie, Lackawanna, Butler, Lebanon, Centre, Somerset, Clearfield, Franklin, Lawrence, Susquehanna, Perry, Monroe, Tioga, Cambria, Jefferson, Mifflin, Venango, Blair, Bradford, Armstrong, Mercer, Fulton, Columbia, Wyoming, Clinton, Crawford, Fayette, Adams, Clarion, Lycoming, Potter, Bedford, Indiana, Huntingdon, McKean, Juniata, Wayne, Pike

**Justification:**
Low margins of error indicate reduced risk in algorithmic bias

2. **Counties requiring additional oversight:** 
Sullivan, Union, Montour, Elk, Greene, Cameron, Snyder, Carbon, Warren

**Justification:**
Although these counties fall in the moderate confidence range, outcomes should be monitored to ensure appropriateness.

3. **Counties needing alternative approaches:** 
Forest

**Justifications:**
Although Forest county falls in the moderate confidence range, it borders the low confidence range and likely should avoid automated decisions until the quality of data improves. 

## Questions for Further Investigation

1. How small or strict should data reliability ranges be?
2. Would it be statistically meaningful to compare margin of error percentages of different variables?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on 09/26/2025

**Reproducibility:** 
- All analysis conducted in R version R 4.5.1
- Census API key required for replication
- Complete code and documentation available at: 
https://musa-5080-fall-2025.github.io/portfolio-setup-aruth3/

**Methodology Notes:**
County selection was determined base on numerical values of reliability rather than categorical indications of reliability due to a lack of categorical variation.

**Limitations:**
Certain population percentage estimates returned values of infinity due to division by denominators with a value of zero. These results should be interpreted with caution as they do not represent meaningful statistical outcomes.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`